---
title: "A Short Instructional PCA Analysis"
author: "Alaric Dinh"
format: html #revealjs for slides 
editor: visual  
prefer-html: true
---

## Introduction

This document will outline my experience in learning how to create a PCA and the resulting data visualizations. We will briefly introduce, in a non rigorous fashion, what a PCA is, the motivation that drives researchers to utilize one, and the basic types of visualizations that can be utilized along with the code utilized to create them.

First, let's develop some intuition about what PCA is. PCA, or principal component analysis is a linear dimensionality reduction technique that is frequently used in exploratory data analysis. Just as statistics (functions of data) allow us to make sense of large datasets through summarization, while mantaining key characteristics of the dataset, PCA allows researchers to analyze multidimensional datasets by reducing the dimensions into a set of principal components that can be represented in a two-dimensional medium.

The mathematical principles at play are beyond the scope of this paper, but the basic process will be outlined and resources for those interested in the mathematics will be in the references. PCA can be considered as being composed of five major steps. The first step, normalization, normalizes the data to a standard normal with mean zero and variance one. Normalization is required when the variables of interest are on a different scale than one another; normalization ensures that we can reasonably compare contributions to variance by ensuring a standard scale.

The second step is the creation of a covariation matrix, which contains the covariance between each variable of interest. The covariation matrix will be used in a process known as "eigendecomposition" which will generate the eigenvectors which will be used to generate the principal components. The eigenvector with the greatest eigenvalue will be in the direction of the greatest variation, and will be considered as the first principal component, the eigenvector with the second greatest eigenvalue will be orthogonal to the first and in the direction of the greatest amount of remaining variance and so on until all principal components are accounted for. There can be as many principal components as there variables of interest.

## The Dataset

Now let's look at how a PCA can be applied to a real dataset (see references for dataset source). PCA can often be useful in agricultural or other biological/ecological settings due to the high number of correlated variables. It is impossible to visualize a dataset with more than three dimensions, so PCA can allow us to mantain the key characteristics of the data while displaying them in two dimensions. let's first load up the required packages tidyverse and factoextra. Then we will import our data and select for only the variables of interest.

```{r}
#| message: false
#First load the dataset (see references for source and information about dataset)
Dataset_Maize_Cordoba <- read.csv("Dataset_Maize_Cordoba.csv")
library(tidyverse)
library(viridis)
library(factoextra)
library(plotly)
library(corrplot)
library(rsconnect)
#Now let's filter out any variables that won't be utilized in the analysis 
DatasetPCA <- Dataset_Maize_Cordoba |> select(Total_N, Total_P, 
                                              Total_K, pH,
                                              Efective_Depth,
                                              Plant_Density_20_days,
                                              Sowing_Seeds_Number,
                                             Yield)
#Removing Spaces 
DatasetPCA2 <- DatasetPCA |> 
  rename(`Total N` = Total_N,
         `Total P` = Total_P,
         `Total K` = Total_K,
         `Effective Depth` = Efective_Depth,
         `Plant Density` = Plant_Density_20_days,
         `Number Sown Seeds` = Sowing_Seeds_Number)


```

Our dataset is comprised of 804 maize cropping events in Cordoba, Colombia from the years 2013-2016. The data was collected by the International Center for Tropical Agriculture (CIAT), the Colombian National Cereals, Legumes Federation (FENALCE), and the Colombian Ministry of Agriculture and Rural Development (MADR) in an effort to better understand how to improve crop yields.

While the dataset contains abundant variables to be analyzed, the author has honed in on particular variables of agronomic interest for this analysis. Our variables of interest are: pH, planting density, Nitrogen levels, Potassium levels, Phosphorus levels, number of seeds sown, and the most important variable yield.

# The Code

With our data imported to R we can now generate our principle components using the prcomp function. The prcomp function takes our dataset, and performs the eigenvalue decomposition creating our orthogonal principal components. Of particular importance is the, scale argument, which performs the normalization step. As our variables are on different scales, it's essential that the scale argument be set to true.

```{r}
#first generate the Principal components 
res.pca2 <- prcomp(DatasetPCA2, 
                  scale = TRUE) #scales to std normal
```

With our Principal Components generated, we can now analyze the principal components with what is known as a scree plot.

```{r}
#visualize the eigenvalues with a scree plot
fviz_screeplot(res.pca2, 
               #adds labels above each component 
               addlabels = TRUE, ylim = c(0, 30))
```

Now that we have our scree plot, we have to exercise some acumen as we are now responsible for choosing which of the principal components we will use to visualize our data. The Y-axis here essentially tells us the amount of variation that is captured by each of these principal components. We can see an inflection point at the 3rd dimension, which tells us that these three principal components are the most significant and should be utilized in our analysis.

There are several ways that we can visualize the data now that we have our principal components. First we will plot the individual observations, where more similar observations are grouped closer together. It should be noted that the functions we will be using to generate our visualizations, will by default utilize the first two principal components unless otherwise specified.

### Individuals plots

```{r}
#creating our color palette with the viridis package
viridis_colors <- viridis(100, end = 0.85)
#create our individuals PCA 
fviz_pca_ind(
  res.pca2, #pass in our PCA generated above
  col.ind = "cos2", #color by representation
  gradient.cols = viridis_colors, #use our Viridis color palette
  label = "none" #remove individual labels
)
```

Just to see how our choice of principal components impacts the visualization let's generate another individual plot using three principal components. The original package used is based upon ggplot2, and therefore we will need to extract the principal components of interest and use an additional package, plotly, to allow us to generate a plot in 3-dimensions.

This plot will just be utilized for demonstration purposes, as it is typically recommended to first generate two dimensional plots using the third principal component to first see if there is any degree of spatial clustering before generating a three dimensional visualization, and academic journals typically will not accept dynamic plots.

```{r}
#Our first three principal components
prin_comp <- prcomp(DatasetPCA2, rank. = 3)

#Extract our principal components 
components <- prin_comp[['x']]

#Convert our components variables to dataframe 
components <- data.frame(components) 

#Create our figure
fig <- plot_ly(components, x = ~PC1, y = ~PC2, z = ~PC3) |> 
  add_markers(size = 12) |> 
  layout(
    title = "Individuals 3D PCA Scatterplot"
  )

fig 

```

While these individuals plots may be of interest in certain situations, with our data we are more interested in the relationships between the variables as we want to know what variables are interacting and impacting yields. Therefore, an alternative visualization known as a variable PCA may be appropriate. Let's create one below.

### Variable PCA

```{r}
#Extract variables 
var <- get_pca_var(res.pca2)
#Now we can visualize the relationships between variables
fviz_pca_var(res.pca2, col.var = "black",
             #prevents labels from overlapping
             repel = TRUE)
```

We can add additional information to this plot by coloring their variables by their contributions to the observed variance.

```{r}
fviz_pca_var(res.pca2,
             #Color by contribution to variance
             col.var = "contrib",
             gradient.cols = viridis_colors,
             repel = TRUE) 
```

An additional visualization, called the PCA biplot, combines elements of the previous two visualizations, displaying both the individuals and variables relationship in one plot. We will generate such a plot now. Before generating the plot, it should be noted that due to the high number of observations (planting events) that the resulting plot may be difficult to interpret, and therefore not the optimal visualization for this dataset.

### Biplot

```{r}
#Generating custom color pallete for biplot
viridis_colors2 <- viridis(20, end = 0.85)
#Generating the biplot 
fviz_pca_biplot(res.pca2, repel = TRUE, label = "var",
                col.var = "contrib",
                gradient.cols = viridis_colors2,
                #we can improve visibility by reducing ind alpha
                alpha.ind = 0.18
                )
```

One final option for visualization is an individual plot, grouped by color. This can be useful when interested in analyzing particular groups, for instance in ecology it may be common to group by a particular taxonomic rank. For our dataset perhaps we want to look at cropping events with similar pH, so let's create a visualization that displays individuals grouped and colored by pH.

### Grouping

```{r}
#Creating our color palette
viridispH <- viridis(7)
fviz_pca_ind(res.pca2,
             label = "none", #remove indvidiual labels
             habillage = DatasetPCA$pH, #color by pH
             palette = viridispH,
             addEllipses = TRUE) 
```

Now that we have looked at the major kinds of standard PCA visualizations, let's look at some useful functions and ways that we can customize these plots. The factoextra package is built upon ggplot2 so any ggplot2 layers can be added with ease. Let's create an example now, and edit our legend title in the previous graph. The formatting is slightly different as instead of adding an additional layer we will simply call the function within the fviz function.

```{r}
#Creating our color palette
  viridispH <- viridis(7)
  fviz_pca_ind(res.pca2,
               label = "none", #remove indvidiual labels
               habillage = DatasetPCA$pH, #color by pH
               palette = viridispH,
               addEllipses = TRUE,
               #change legend title
               legend.title = "pH",
               #change title to bold 
               ggtheme = theme(
                 plot.title = element_text(face = "bold", size = 16)
               ))
```

# Other Functions

Although not explored in depth in this document, we have listed some useful functions that can be found in the factoextra package, documentation for the package can be found here <https://cran.r-project.org/web/packages/factoextra/refman/factoextra.html>

1.  fviz_ca \| Graphing correspondence analysis

2.  fviz_ma \| Graphing multifactor analysis

3.  fviz_cos2 \| Visualize quality of representation from PCA/CA/MCA analysis

4.  facto_summarize \| Summarize the outputs of factor analysis

5.  get_pca \| Extract all results from PCA

    ### Variable Correlation Plots

let's look at one final useful technique when performing PCA, analysis of the variables' contribution to the observed variance. Another useful package here is the corrplot package, which can enable us to create some helpful visualizations.

```{r}
#create our correlation plot
corrplot(var$contrib,
         #is a correlation matrix
         is.corr = FALSE)
```

We can also create barplots to assess variables' contributions to particular components. The axes argument here is critical as it allows us to select which principal components we wish to analyze. For instance we may be interested in looking at pH and Total N's contribution to component one.

### Component Variable Contribution Barplots

```{r}
fviz_contrib(res.pca2, 
             #contribution to variance plot
             choice = "var",
             #axes selected
             axes = 1)
```

# Tips & Interpretation

### Eigenvalue & Component Selection

There are several schools of thought when selecting principal components, one common method is to select an arbitrary percentage of variance to be represented by the principal components such as 50 percent.

The other approach is to analyze components that have specific variables of interest, for instance I may opt to select component three as one of my axes as I'm most interested in how Phosphorus interacts with other variables of interest.

### Individuals

Individual plots are primarily used for spotting spatial clustering of data, its important to remember that each component explains a differing amount of variation in the data so horizontal spread is more impactful than vertical spread, and the percentages of each component should be kept in mind.

Remember to check the contribution plot when analyzing clusters of data, as if clustering occurs along an axis its important to check the contribution plot to assess what variables are responsible for the observed differences between points.

### Biplots

There are several aspects of the biplot that can be assessed, vector's magnitudes and relative directions. The longer vectors are, the more that particular variable contributes to the observed variance. The second thing to assess is the variable's relationship to one another. Vectors pointing in the same direction are correlated to a greater degree, for instance in our dataset we can see that Total Phosphorus and Yield are greatly correlated.

Biplots are useful for assessing which variables contribute to variance in your sample. In our dataset an interesting result is that Yield and pH are almost independent, with the resulting vectors being close to orthogonal. This highlights the importance of incorporating expertise and our knowledge of the data source, as it is well known that pH greatly impacts yield, however, in this sample this doesn't seem to be the case. This is likely due to farmers only planting crops in soils with optimal pH ranges, which can easily be assessed by checking the range of the pH variable.

### Useful tips

The viridis color pallete while colorblind friendly, can be difficult to read at extreme values. The "end" argument can prevent extremely bright yellows from being used allowing for better visibility.

Factoextra is built upon ggplot so ggplot elements can be added within the desired function, or the plot can be saved and additional layers added using the "+" syntax.

# Conclusion

Though it is easy to create some visualizations with a package such as factoextra, it does not mean that it should be done. PCA is only a dimensionality reduction technique, and its chief purpose is to minimize the loss of dispersion while simplifying the dataset. It is important to understand what the results mean, and what conclusions can and can't be drawn from the data. PCA is an exploratory method and should not be used to make inferences.

# References

1.  Gewers, F. L., Ferreira, G. R., Arruda, H. F. D., Silva, F. N., Comin, C. H., Amancio, D. R., & Costa, L. D. F. (2021). Principal Component Analysis: A Natural Approach to Data Exploration. *ACM Computing Surveys (CSUR)*, *54*(4), 70:1-70:34. <https://doi.org/10.1145/3447755>

2.  Jimenez, D., Delerce, S. J., Dorado, H. A., Cock, J., Muñoz, L. A., Agamez, A., & Jarvis, A. (2019). *Cropping events of maize in Cordoba-Colombia* \[Dataset\]. Harvard Dataverse. <https://doi.org/10.7910/DVN/LSDQ5C>

3.  *Mathematical Approach to PCA*. (11:38:29+00:00). GeeksforGeeks. <https://www.geeksforgeeks.org/machine-learning/mathematical-approach-to-pca/>

4.  *Pca*. (n.d.). Retrieved May 27, 2026, from <https://plotly.com/r/pca-visualization/>

5.  *PCA in R: A Step-by-Step Tutorial with Examples*. (n.d.). Retrieved May 27, 2026, from <https://www.datacamp.com/tutorial/pca-analysis-r>

6.  PhD, S. L. (2020, May 30). PCA clearly explained - How, when, why to use it and feature importance: A guide in Python. *Towards Data Science*. <https://towardsdatascience.com/pca-clearly-explained-how-when-why-to-use-it-and-feature-importance-a-guide-in-python-7c274582c37e/>

7.  *Visualization of PCA in R: Scatter & Trejactory \| Visual*. (n.d.). Retrieved May 27, 2026, from <https://yzwisalaity.github.io/PCA-Visualization/>
